Arrows for Parallel Computation

### 2019-12-27 paper/content/ms.tex
shortening; explaining speedup

\ptcomment{The non-strict semantics of Haskell, and the fact that reduction encapsulates computations as closures, makes it relatively easy to define alternate parallelisations. A range of approaches have been explored, including data parallelism (Chakravarty et al., 2007; Keller et al.,
2010), GPU-based approaches (Mainland & Morrisett, 2010; Svensson, 2011), software
transactional memory (Harris et al., 2005; Perfumo et al., 2008). The Haskell–GPU bridge
Accelerate (Chakravarty et al., 2011; Clifton-Everest et al., 2014; McDonell et al., 2015)
deserves a special mention. <<Why does it deserve special mention?>>  Accelerate is completely orthogonal to our approach. A good survey of parallel Haskells can be found in ~\cite{Marlow2013}.

Our PArrow implementation uses three task parallel languages as backends: the
GpH (Trinder et al., 1996, 1998) parallel Haskell dialect and its multicore version (Marlow
et al., 2009), the \ensuremath{\Conid{Par}} Monad (Marlow et al., 2011; Foltzer et al., 2012), and Eden (Loogen
et al., 2005; Loogen, 2012). These languages are under active development, for example a combined shared and distributed memory implementation of GpH is available (Aljabri et al., 2014,
2015). Research on Eden includes a low-level  implementation (Berthold, 2008; Berthold et al., 2016), work on skeleton
composition (Dieterle et al., 2016), communication (Dieterle et al., 2010a), and generation
of process networks (Horstmeyer & Loogen, 2013). The definitions of new Eden skeletons is a specific focus (Hammond et al., 2003; Berthold
& Loogen, 2006; Berthold et al., 2009b,c; Dieterle et al., 2010b; de la Encina et al., 2011;
Dieterle et al., 2013; Janjic et al., 2013).

Other task parallel Haskells, related to Eden, GpH, and the \ensuremath{\Conid{Par}} Monad, include the following.  
HdpH (Maier et al., 2014; Stewart et al., 2016) is an extension of \ensuremath{\Conid{Par}} Monad to
heterogeneous clusters. LVish (Kuper et al., 2014) is a communication-centred extension
of \ensuremath{\Conid{Par}} Monad.}

@@

\paragraph{Arrows.}
Arrows were introduced by \citet{HughesArrows} as a less restrictive alternative to Monads, in essence they are a generalised function arrow~\ensuremath{\to }. \citet{Hughes2005} presents a tutorial on Arrows. \citet{jacobs_heunen_hasuo_2009,LINDLEY201197,ATKEY201119} develop theoretical background of Arrows. \citet{Paterson:2001:NNA:507669.507664} introduced a new notation for Arrows. Arrows have applications in information flow research \cite{1648705,LI20101974,Russo:2008:LLI:1411286.1411289}, invertible programming \cite{Alimarine:2005:BAA:1088348.1088357}, and quantum computer simulation \cite{vizzotto_altenkirch_sabry_2006}. But probably most prominent application of Arrows is Arrow-based functional reactive programming, AFRP \cite{Nilsson:2002:FRP:581690.581695,Hudak2003,Czaplicki:2013:AFR:2499370.2462161}.
\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of Arrows that capsule the computation more than regular Arrows do and thus enable optimisations. Their approach would allow parallel composition, as their special Arrows would not interfere with each other in concurrent execution. In contrast, we capture a whole parallel computation as a single entity: our main instantiation function \ensuremath{\Varid{parEvalN}} makes a single (parallel) Arrow out of list of Arrows. \citet{Huang2007} utilise Arrows for parallelism, but strikingly different from our approach. They use Arrows to orchestrate several tasks in robotics. We, however, propose a general interface for parallel programming, while remaining completely in Haskell.

\paragraph{Arrows in other languages.}
Although this work is centred on Haskell implementation of Arrows, it is applicable to any functional programming language where parallel evaluation and Arrows can be defined. Basic definitions of PArrows are possible in the Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM). However, they are beyond the scope of this work, as are similar experiments with the Eta language\footnote{Eta project page at \url{http://eta-lang.org}}, 
a new approach to Haskell on the JVM.

@@

While we could have based our DSL on Profunctors as well, we chose Arrows\mbcomment{cite missing?} for this paper since they they allow for a more direct way of thinking about parallelism than general Profunctors because of their composability. However, they are a promising candidate for future improvements of our DSL. Some Profunctors, especially ones supporting a composition operation, choice, and looping, can already be adapted to our interface as shown in Appendix~\ref{app:profunctorArrows}.
	\subsection{Short introduction to parallel Haskells}
\label{sec:parallelHaskells}
\label{sec:parEvalNIntro}
In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \ensuremath{\Varid{a}\to \Varid{b}} in parallel or \ensuremath{\Varid{parEvalN}\mathbin{::}[\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]}, as also Figure~\ref{fig:parEvalN} symbolically shows.
\begin{figure}[t]
  \centering
	\includegraphics[scale=0.7]{parEvalN}
	\caption{Schematic illustration of \ensuremath{\Varid{parEvalN}}. A list of inputs is transformed by different functions in parallel.}
	\label{fig:parEvalN}
\end{figure}

In this section, we will implement this non-Arrow version which will later be adapted for usage in our Arrow-based parallel Haskell.

There exist several parallel Haskells already. Among the most important are probably GpH \citep[based on \ensuremath{\Varid{par}} and \ensuremath{\Varid{pseq}} \enquote{hints},][]{Trinder1996,Trinder1998a}, the \ensuremath{\Conid{Par}} Monad \citep[a monad for deterministic parallelism,][]{par-monad,Foltzer:2012:MPC:2398856.2364562}, Eden \citep[a parallel Haskell for distributed memory,][]{eden,Loogen2012}, HdpH \citep[a Template Haskell-based parallel Haskell for distributed memory,][]{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} and LVish \citep[a \ensuremath{\Conid{Par}} extension with focus on communication,][]{Kuper:2014:TPE:2666356.2594312}.

As the goal of this paper is not to re-implement yet another parallel runtime, but to represent parallelism with Arrows, we base our efforts on existing work which we wrap as backends behind a common interface. For this paper we chose GpH for its simplicity, the \ensuremath{\Conid{Par}} Monad to represent a monadic DSL, and Eden as a distributed parallel Haskell.

LVish and HdpH were not chosen as the former does not differ from the original \ensuremath{\Conid{Par}} Monad with regard to how we would have used it in this paper, while the latter (at least in its current form) does not comply with our representation of parallelism due to its heavy reliance on Template Haskell.

We will now go into some detail on GpH, the \ensuremath{\Conid{Par}} Monad and Eden, and also give their respective implementations of the non-Arrow version of \ensuremath{\Varid{parEvalN}}.

%Before we go into detail on how we can use this idea of parallelism
%for parallel Arrows, as a short introduction to parallelism in
%Haskell we will now implement |parEvalN| with several different
%parallel Haskells.

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{parEvalN}\mathrel{=}\Varid{spawnF}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% \begin{figure}[h]
% 	\includegraphics[scale=0.5]{images/parEvalNEden}
% 	\caption{Dataflow of the Eden |parEvalN| version.}
% 	\label{fig:parEvalNEden}
% \end{figure}
% A simplistic graphical depiction of this definition can be found in Fig.~\ref{fig:parEvalNEden}.

\paragraph{Eden TraceViewer.}
\label{sec:edentv}
To comprehend the efficiency and the lack thereof in a parallel program, an inspection of its execution is extremely helpful. While some large-scale solutions exist \cite{Geimer2010}, the parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See \url{http://hackage.haskell.org/package/edentv} on Hackage for the last available version of Eden TraceViewer.} \cite{Berthold2007a}. In the next sections we will present some \emph{trace visualisations}, the post-mortem process diagrams of Eden processes and their activity.

@@

In our communication example we can use this \ensuremath{\Conid{Future}} concept for direct communication between nodes as shown in Fig.~\ref{fig:someCombinatorParallel}.
\begin{figure}[t]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{someCombinator}\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;(\Varid{fut}\;\Varid{b})\;(),{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{fut}\;\Varid{b})\;\Varid{c}\;(),{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{Future}\;\Varid{fut}\;\Varid{b}\;())\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{arr}\;\Varid{b}\;\Varid{c}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{c}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{someCombinator}\;\Varid{fs1}\;\Varid{fs2}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{parEvalN}\;()\;(\Varid{map}\;(\mathbin{>\!\!>\!\!>}\Varid{put}\;())\;\Varid{fs1})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{rightRotate}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{parEvalN}\;()\;(\Varid{map}\;(\Varid{get}\;()\mathbin{>\!\!>\!\!>})\;\Varid{fs2}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{The outline combinator in parallel.}
\label{fig:someCombinatorParallel}
\end{figure}
In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed -- similar to what is shown in the trace visualisation in Fig.~\ref{fig:withFutures}. One especially elegant aspect of the definition in Fig.~\ref{fig:future} is that we can specify the type of \ensuremath{\Conid{Future}} to be used per backend with full interoperability between code using different backends, without even requiring to know about the actual type used for communication. We only specify that there has to be a compatible Future and do not care about any specifics as can be seen in Fig.~\ref{fig:someCombinatorParallel}. With the PArrows DSL we can also define default instances \ensuremath{\Conid{Future}\;\Varid{fut}\;\Varid{a}\;()} for each backend similar to how \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;()} was defined in Section~\ref{sec:parallel-arrows}. Details can be found in Section~\ref{app:omitted}. \olcomment{Fig. is not really clear. Do Figs with a lot of load? --- fft?}
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{withFutures}
	\caption[with Futures]{Communication between 4 Eden processes with Futures. Other than in Fig.~\ref{fig:withoutFutures}, processes communicate directly (one example message is highlighted) instead of always going through the master node (bottom bar).}
	\label{fig:withFutures}
\end{figure}
	%\pagebreak
	\section{Skeletons}
\label{sec:skeletons}
Now we have developed Parallel Arrows far enough to define some useful algorithmic skeletons that abstract typical parallel computations. While there are many possible skeletons to implement, we demonstrate the expressive power of PArrows here using four \ensuremath{\Varid{map}}-based and three toplogical skeletons.
%%% \FloatBarrier
\subsection{\ensuremath{\Varid{map}}-based skeletons}
\label{sec:map-skeletons}
The essential differences between the mapping skeletons presented here are in terms of order of evaluation and work distribution but still provide the same semantics as a sequential \ensuremath{\Varid{map}}.

@@

If we used the original definition of \ensuremath{\Varid{parEvalN}}, however, these skeletons would produce an infinite loop with the GpH and \ensuremath{\Conid{Par}} Monad which during runtime would result in the program crash. This materialises with the usage of \ensuremath{\Varid{loop}} of the \ensuremath{\Conid{ArrowLoop}} type class and we think that this is due to difference of how evaluation is done in these backends when compared to Eden. An investigation of why this difference exists is beyond the scope of this work, we only provide a workaround for these types of skeletons as such they probably are not of much importance outside of a distributed memory environment. However our workaround enables users of the DSL to test their code within a shared memory setting.
% even though other skeletons would be better suited to be run with them. --- OL: I don't get this sentence part
The idea of the fix is to provide a \ensuremath{\Conid{ArrowLoopParallel}} type class that has two functions -- \ensuremath{\Varid{loopParEvalN}} and \ensuremath{\Varid{postLoopParEvalN}}, where the first is to be used inside an \ensuremath{\Varid{loop}} construct while the latter will be used right outside of the \ensuremath{\Varid{loop}}. This way we can delegate to the actual \ensuremath{\Varid{parEvalN}} in the spot where the backend supports it.
\>[B]{}\mathbf{class}\;\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}\Rightarrow {}\<[E]%
\>[9]{}\Conid{ArrowLoopParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}\;\mathbf{where}{}\<[E]%
\>[5]{}\Varid{loopParEvalN}\mathbin{::}\Varid{conf}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\>[5]{}\Varid{postLoopParEvalN}\mathbin{::}\Varid{conf}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
As Eden has no problems with the looping skeletons, we use this instance:
\>[B]{}\mathbf{instance}\;(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Conid{Conf})\Rightarrow {}\<[E]%
\>[9]{}\Conid{ArrowLoopParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Conid{Conf}\;\mathbf{where}{}\<[E]%
\>[5]{}\Varid{loopParEvalN}\mathrel{=}\Varid{parEvalN}{}\<[E]%
\>[5]{}\Varid{postLoopParEvalN}\;\anonymous \mathrel{=}\Varid{evalN}{}\<[E]%
As \ensuremath{\Conid{Par}} Monad and GpH have problems with \ensuremath{\Varid{parEvalN}} inside of \ensuremath{\Varid{loop}} their respective instances for \ensuremath{\Conid{ArrowLoopParallel}} look like this:
\>[B]{}\mathbf{instance}\;(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;(\Conid{Conf}\;\Varid{b}))\Rightarrow {}\<[E]%
\>[9]{}\Conid{ArrowLoopParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;(\Conid{Conf}\;\Varid{b})\;\mathbf{where}{}\<[E]%
\>[5]{}\Varid{loopParEvalN}\;\anonymous \mathrel{=}\Varid{evalN}{}\<[E]%
\>[5]{}\Varid{postLoopParEvalN}\mathrel{=}\Varid{parEvalN}{}\<[E]%

@@

The \ensuremath{\Conid{Par}} monad and GpH -- in its multicore version \cite{Marlow2009} --  can be executed on shared memory machines only.
Although GpH is available on distributed memory
clusters, and newer distributed memory Haskells such as HdpH exist,
current support of distributed memory in PArrows is limited to
Eden. We used the MPI backend of Eden in a distributed memory
setting. However, for shared memory Eden features a \enquote{CP} backend
that merely copies the memory blocks between disjoint heaps. In
this mode, Eden still operates in the \enquote{nothing shared} setting, but
is adapted better to multicore machines. We call this version of Eden
\enquote{Eden~CP}.

@@

We compare the PArrow performance with direct implementations of the benchmarks in Eden, GpH and the \ensuremath{\Conid{Par}} Monad.
We start with the definition of mean overhead to compare both PArrows-enabled and standard benchmark implementations. We continue comparing speedups and overheads for the shared memory implementations and then study OpenMPI variants of the Eden-enabled PArrows as a representative of a distributed memory backend. We plot all speedup curves and all overhead values in the supplementary materials.

\subsubsection{Defining overhead}

@@

\subsubsection{Shared memory}

>>>>

@@

\paragraph{Arrows.}
Arrows were introduced by \citet{HughesArrows} as a less restrictive alternative to Monads, in essence they are a generalised function arrow~\ensuremath{\to }. \citet{Hughes2005} presents a tutorial on Arrows. \citet{jacobs_heunen_hasuo_2009,LINDLEY201197,ATKEY201119} develop theoretical background of Arrows. \citet{Paterson:2001:NNA:507669.507664} introduced a new notation for Arrows. Arrows have applications in information flow research \cite{1648705,LI20101974,Russo:2008:LLI:1411286.1411289}, invertible programming \cite{Alimarine:2005:BAA:1088348.1088357}, and quantum computer simulation \cite{vizzotto_altenkirch_sabry_2006}. But probably most prominent application of Arrows is Arrow-based functional reactive programming, AFRP \cite{Nilsson:2002:FRP:581690.581695,Hudak2003,Czaplicki:2013:AFR:2499370.2462161}.
\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of Arrows that capsule the computation more than regular Arrows do and thus enable optimisations. Their approach would allow parallel composition, as their special Arrows would not interfere with each other in concurrent execution. In contrast, we capture a whole parallel computation as a single entity: our main instantiation function \ensuremath{\Varid{parEvalN}} makes a single (parallel) Arrow out of a list of Arrows. \citet{Huang2007} utilise Arrows for parallelism, but strikingly different from our approach. They use Arrows to orchestrate several tasks in robotics. We, however, propose a general interface for parallel programming, while remaining completely in Haskell.
\citet{zhang_session_2019} introduced a session-type framework that generates parallel code. As they stated, their approach to express dataflow with Arrows was inspired by our formalism.

\paragraph{Arrows in other languages.}
Although this work is centred on Haskell implementation of Arrows, it is applicable to any functional programming language where parallel evaluation and Arrows can be defined. Preliminary experiments show that basic definitions of PArrows are possible in the Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM). However, they are beyond the scope of this work, as are similar experiments with the Eta language\footnote{Eta project page at \url{http://eta-lang.org}}, 
a new approach to Haskell on the JVM.

@@

While our DSL could have been based on Profunctors \citep[see, e.g.][]{boisseau_what_2018} directly, we chose Arrows for this paper for the easier representation of composability. Special Profunctors supporting a composition operation, choice, and looping, can already be used as a foundation of our interface as shown in Appendix~\ref{app:profunctorArrows}.
	\subsection{Short introduction to parallel Haskells}
\label{sec:parallelHaskells}
\label{sec:parEvalNIntro}
In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \ensuremath{\Varid{a}\to \Varid{b}} in parallel or \ensuremath{\Varid{parEvalN}\mathbin{::}[\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]}, as also Figure~\ref{fig:parEvalN} symbolically shows.
\begin{figure}[t]
  \centering
	\includegraphics[scale=0.7]{parEvalN}
	\caption{Schematic illustration of \ensuremath{\Varid{parEvalN}}. A list of inputs is transformed by different functions in parallel.}
	\label{fig:parEvalN}
\end{figure}

In this section, we will implement this non-Arrow version which will later be adapted for usage in our Arrow-based parallel Haskell.
As the goal of this paper is not to re-implement yet another parallel runtime, but to represent parallelism with Arrows, we base our efforts on existing work which we wrap as backends behind a common interface. For this paper we chose GpH for its simplicity, the \ensuremath{\Conid{Par}} Monad to represent a monadic DSL, and Eden as a distributed parallel Haskell.

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{parEvalN}\mathrel{=}\Varid{spawnF}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% \begin{figure}[h]
% 	\includegraphics[scale=0.5]{parEvalNEden}
% 	\caption{Dataflow of the Eden |parEvalN| version.}
% 	\label{fig:parEvalNEden}
% \end{figure}
% A simplistic graphical depiction of this definition can be found in Fig.~\ref{fig:parEvalNEden}.

\paragraph{Eden TraceViewer.}
\label{sec:edentv}
To comprehend the efficiency or the lack thereof in a parallel program, an inspection of its execution is extremely helpful. While some large-scale solutions exist \cite{Geimer2010}, the parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See \url{http://hackage.haskell.org/package/edentv} on Hackage for the last available version of Eden TraceViewer.} \cite{Berthold2007a}. In the next sections we will present some \emph{trace visualisations}, the post-mortem process diagrams of Eden processes and their activity.

@@

In our communication example we can use this \ensuremath{\Conid{Future}} concept for direct communication between nodes as shown in Fig.~\ref{fig:someCombinatorParallel}.
\begin{figure}[t]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{someCombinator}\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;(\Varid{fut}\;\Varid{b})\;(),{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{fut}\;\Varid{b})\;\Varid{c}\;(),{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{Future}\;\Varid{fut}\;\Varid{b}\;())\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{arr}\;\Varid{b}\;\Varid{c}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{c}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{someCombinator}\;\Varid{fs1}\;\Varid{fs2}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{parEvalN}\;()\;(\Varid{map}\;(\mathbin{>\!\!>\!\!>}\Varid{put}\;())\;\Varid{fs1})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{rightRotate}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{parEvalN}\;()\;(\Varid{map}\;(\Varid{get}\;()\mathbin{>\!\!>\!\!>})\;\Varid{fs2}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{The outline combinator in parallel.}
\label{fig:someCombinatorParallel}
\end{figure}
In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed---similar to what is shown in the trace visualisation in Fig.~\ref{fig:withFutures}. One especially elegant aspect of the definition in Fig.~\ref{fig:future} is that we can specify the type of \ensuremath{\Conid{Future}} to be used per backend with full interoperability between code using different backends, without even requiring to know about the actual type used for communication. We only specify that there has to be a compatible Future and do not care about any specifics as can be seen in Fig.~\ref{fig:someCombinatorParallel}. With the PArrows DSL we can also define default instances \ensuremath{\Conid{Future}\;\Varid{fut}\;\Varid{a}\;()} for each backend similar to how \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;()} was defined in Section~\ref{sec:parallel-arrows}. Details can be found in Section~\ref{app:omitted}. \olcomment{Fig. is not really clear. Do Figs with a lot of load? --- fft?}
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{withFutures}
	\caption[with Futures]{Communication between 4 Eden processes with Futures. Other than in Fig.~\ref{fig:withoutFutures}, processes communicate directly (one example message is highlighted) instead of always going through the master node (bottom bar).}
	\label{fig:withFutures}
\end{figure}
	%\pagebreak
	\section{Skeletons}
\label{sec:skeletons}
Now we have developed Parallel Arrows far enough to define some useful algorithmic skeletons that abstract typical parallel computations. While there are many possible skeletons to implement, we demonstrate the expressive power of PArrows here using four \ensuremath{\Varid{map}}-based and three topological skeletons.
%%% \FloatBarrier
\subsection{\ensuremath{\Varid{map}}-based skeletons}
\label{sec:map-skeletons}
The essential differences between the mapping skeletons presented here are in terms of order of evaluation and work distribution but still provide the same semantics as a sequential \ensuremath{\Varid{map}}.

@@

If we used the original definition of \ensuremath{\Varid{parEvalN}}, however, these skeletons would produce an infinite loop with the GpH and \ensuremath{\Conid{Par}} Monad which during runtime would result in the program crash. This materialises with the usage of \ensuremath{\Varid{loop}} of the \ensuremath{\Conid{ArrowLoop}} type class and we think that this is due to difference of how evaluation is done in these backends when compared to Eden. We provide a workaround for these types of skeletons as such they probably are not of much importance outside of a distributed memory environment. However our workaround enables users of the DSL to test their code within a shared memory setting.
% even though other skeletons would be better suited to be run with them. --- OL: I don't get this sentence part
The idea of the fix is to provide a \ensuremath{\Conid{ArrowLoopParallel}} type class that has two functions---\ensuremath{\Varid{loopParEvalN}} and \ensuremath{\Varid{postLoopParEvalN}}, where the first is to be used inside an \ensuremath{\Varid{loop}} construct while the latter will be used right outside of the \ensuremath{\Varid{loop}}. This way we can delegate to the actual \ensuremath{\Varid{parEvalN}} in the spot where the backend supports it.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowLoopParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{loopParEvalN}\mathbin{::}\Varid{conf}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{postLoopParEvalN}\mathbin{::}\Varid{conf}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
As Eden has no problems with the looping skeletons, we use this instance:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Conid{Conf})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowLoopParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Conid{Conf}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{loopParEvalN}\mathrel{=}\Varid{parEvalN}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{postLoopParEvalN}\;\anonymous \mathrel{=}\Varid{evalN}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
As \ensuremath{\Conid{Par}} Monad and GpH have problems with \ensuremath{\Varid{parEvalN}} inside of \ensuremath{\Varid{loop}} their respective instances for \ensuremath{\Conid{ArrowLoopParallel}} look like this:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;(\Conid{Conf}\;\Varid{b}))\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowLoopParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;(\Conid{Conf}\;\Varid{b})\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{loopParEvalN}\;\anonymous \mathrel{=}\Varid{evalN}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{postLoopParEvalN}\mathrel{=}\Varid{parEvalN}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

The \ensuremath{\Conid{Par}} monad and GpH---in its multicore version \cite{Marlow2009}--- can be executed on shared memory machines only.
Although GpH is available on distributed memory
clusters, and newer distributed memory Haskells such as HdpH exist,
current support of distributed memory in PArrows is limited to
Eden. We used the MPI backend of Eden in a distributed memory
setting. However, for shared memory Eden features a \enquote{CP} backend
that merely copies the memory blocks between disjoint heaps. In
this mode, Eden still operates in the \enquote{nothing shared} setting, but
is adapted better to multicore machines. We call this version of Eden
\enquote{Eden~CP}.

@@

We compare the PArrow performance with direct implementations of the benchmarks in Eden, GpH and the \ensuremath{\Conid{Par}} Monad.
We start with the definition of mean overhead in Section~\ref{sec:def-overhead}. We continue comparing speedups and overheads for the shared memory implementations (Section~\ref{sec:pref-shm}) and  OpenMPI variants (Section~\ref{sec:pref-dm}). We plot all speedup curves and all overhead values in the supplementary materials.

\subsubsection{Defining overhead and speedup}
\label{sec:def-overhead}

@@

The speedup is a measure of parallel quality. Basically, it is a quotient of sequential execution time to the parallel execution time; speedup measures the speed factor from running a parallel program compared to a sequential one. The ideal speedup is linear speedup, e.g. 32 for 32 cores.

\subsubsection{Shared memory}
\label{sec:pref-shm}
